# How to Use This Book {#sec-how-to-use}

## For Students

### The Exploration Cycle

Each chapter in this book is built around an **exploration**—a guided investigation of statistical concepts through data, code, and questions. The typical cycle is:

```
READ Scenario → RUN Code → ANSWER Questions →
  EXPERIMENT → BREAK Things → FIX Them →
    DISCOVER Patterns → FORMALIZE Concepts →
      DISCUSS → EXTEND
```

### Recommended Workflow

**Step 1: Read the Scenario** (5-10 minutes)

Each chapter starts with a story—often involving fictional characters facing real statistical challenges. Read it carefully. These scenarios:

- Motivate why we care about the statistical concepts
- Frame the questions we'll explore
- Provide context for interpreting results

**Step 2: Set Up Your Computing Environment** (5-15 minutes)

```{r, eval=FALSE}
# Install required packages (first time only)
install.packages(c("tidyverse", "here", "knitr"))

# Load packages
library(tidyverse)
library(here)

# Download data
# (See the book's GitHub repository)
```

**Step 3: Work Through the Exploration** (60-120 minutes)

This is where the real learning happens. For each code chunk:

1. **Read** the code before running it
2. **Predict** what it will do
3. **Run** it and see if you were right
4. **Examine** the output carefully
5. **Modify** the code to test your understanding

For example, if you see:

```{r, eval=FALSE}
mean(data$hours_helping, na.rm = TRUE)
```

Ask yourself:

- What is `mean()` doing?
- What is `na.rm`? Why `TRUE`?
- What if I change it to `FALSE`?
- What if I use `median()` instead?

**Try it!** The best way to understand code is to break it and fix it.

**Step 4: Answer the Questions** (30-60 minutes)

Each exploration poses questions. Some are:

- **Mechanical**: "What does this code do?"
- **Conceptual**: "Why does the trimmed mean differ from the regular mean?"
- **Evaluative**: "When should we use one versus the other?"
- **Creative**: "How would you approach this differently?"

Write out your answers. Writing forces clarity. You might:

- Create an R Markdown or Quarto document
- Keep a statistics notebook
- Type answers in a shared document (if working in a group)

**Step 5: Experiment and Extend** (30-90 minutes)

Once you've worked through the provided code, try:

- **Different data**: Load your own data or find public data
- **Different methods**: Try alternative approaches you read about
- **Different questions**: Pose your own statistical questions
- **Simulations**: Simulate data to test your understanding

The explorations are starting points, not finish lines.

**Step 6: Read and Discuss** (30-60 minutes)

Each chapter includes **recommended readings**. You don't have to read everything, but:

- **Skim** all of them to see which speaks to you
- **Read deeply** 1-2 that clarify things or offer new perspectives
- **Compare** different authors' explanations

Then discuss (with study group, classmates, or just by writing):

- What did you learn from the exploration?
- What's still confusing?
- How does this connect to your research interests?
- What new questions arose?

### Working in Groups

Many students find it helpful to work in groups of 2-4 people. This is encouraged, with some caveats:

**DO**:

- Work through the exploration together, talking through ideas
- Share strategies for debugging code
- Discuss conceptual questions
- Divide up readings and summarize for each other
- Teach each other what you understand

**DON'T**:

- Divide up the work so each person only does part of it
- Just copy someone else's code without understanding it
- Skip parts because "someone else got it"
- Let one person do all the coding while others watch

**The goal** is for everyone in the group to engage with all parts of the exploration.

### When You Get Stuck

#### On Code

**Error messages** are your friends (once you learn to read them):

```{r, eval=FALSE}
mean(hours_helping, na.rm = TRUE)
# Error: object 'hours_helping' not found
```

This tells you: R can't find `hours_helping`. Maybe you need `data$hours_helping` or you haven't loaded the data yet.

**Debugging strategies**:

1. Read the error message carefully
2. Check for typos
3. Verify you've loaded necessary packages
4. Try a simpler version of the code
5. Search the exact error message online
6. Use `?function_name` to read help files
7. Ask an AI assistant (but verify you understand the answer)

#### On Concepts

**Confusion is normal**. Statistics is subtle. When concepts aren't clicking:

1. Try to explain it to someone else (or write it out)
2. Draw a picture or diagram
3. Simulate a simple example where you know the answer
4. Read an alternative explanation
5. Come to office hours with specific questions
6. Sleep on it (seriously—your brain processes while you sleep)

#### On "Why Would I Ever Use This?"

Sometimes you'll learn methods that seem artificial or impractical. Trust that:

- They'll make sense later when we connect ideas
- They're building blocks for more complex methods
- Understanding the simple case helps you understand the complex case

But also: ask! "When would I actually use this?" is a great question.

## For Instructors

### Using This Book in a Flipped Classroom

This book is designed for a **flipped classroom** where students explore before class and come with questions to discuss.

**Typical Weekly Schedule**:

- **Monday**: Release new exploration chapter
- **Tuesday-Thursday**: Students work through exploration (individually or in groups)
- **Friday (or Monday)**: Class discussion

### Class Time Structure

**Option 1: Question-Driven Discussion** (recommended)

1. **Collect questions** (5 min): Students write top questions on board/doc
2. **Prioritize** (5 min): Vote or instructor selects most generative questions
3. **Discuss** (60-75 min): Work through questions, with mini-lectures as needed
4. **Synthesize** (10 min): Summarize key takeaways, preview next week

**Option 2: Structured Activities**

1. **Pair-share** (10 min): Pairs discuss what confused them most
2. **Live coding** (20 min): Instructor demonstrates key techniques
3. **Group problem-solving** (30 min): Groups tackle extension problems
4. **Report out** (10 min): Groups share approaches

**Option 3: Hybrid**

Mix question-driven discussion with planned mini-lectures on key concepts that students often struggle with.

### Assessment

**Explorations** (40%):

- Graded on completeness and thoughtfulness, not correctness
- "Satisfactory" = genuine attempt to answer all questions, run code, experiment
- Can be done in groups

**Individual Quizzes** (20%):

- Short (3-5 questions), focused on key concepts from previous week
- Ensures individual understanding

**Final Project** (30%):

- Replication of published study with new methods
- Pre-analysis plan for own research
- Methods paper investigating a specific technique

**Participation** (10%):

- Asking/answering questions in class
- Helping classmates

### Customization

You can:

- **Reorder** chapters (after Part I) to fit your emphasis
- **Skip** chapters if your students have prior background
- **Supplement** with your own examples/data from your field
- **Extend** explorations with additional questions
- **Create** new explorations using the same format

All materials are open-source. Fork and adapt as needed.

### Support

For teaching materials (slides, answer keys, additional exercises):

- See the book's GitHub repository
- Join the instructors' mailing list
- Contribute your own materials

## For Self-Learners

Using this book on your own? You can succeed, but it requires discipline and creativity.

**Create structure**:

- Set a regular schedule (e.g., one chapter per week)
- Block out time for exploration (treat it like a class meeting)
- Set deadlines for yourself

**Find community**:

- Join or create a study group (virtual or in-person)
- Find an online forum or Slack channel
- Share your work on GitHub or a blog
- Ask questions on Stack Overflow, Cross Validated, or similar

**Get feedback**:

- Post your code and ask for review
- Explain concepts on social media (teaching helps learning)
- Apply methods to your own data and share results

**Supplement**:

- Watch videos of statistics lectures (to hear alternative explanations)
- Read widely (different textbooks, blog posts, papers)
- Do the recommended readings
- Seek out research papers that use methods you're learning

**Check understanding**:

- Can you explain the concept to a non-statistician?
- Can you write code from scratch (not just run provided code)?
- Can you modify methods for new contexts?
- Can you evaluate when methods are appropriate?

If yes to all four, you've mastered it. If not, keep exploring.

## Technical Setup

### Required Software

**R** (version 4.0 or later):

- Download from [https://cran.r-project.org/](https://cran.r-project.org/)
- Free and open-source

**RStudio** (recommended but not required):

- Download from [https://posit.co/downloads/](https://posit.co/downloads/)
- Makes R much easier to use
- Alternatives: VS Code, Jupyter, command line

**Quarto** (if you want to render documents like the book):

- Download from [https://quarto.org/](https://quarto.org/)
- Integrates with RStudio

### Required R Packages

Install with:

```{r, eval=FALSE}
install.packages(c(
  "tidyverse",      # Data manipulation and visualization
  "here",           # File path management
  "knitr",          # Document generation
  "broom",          # Tidy model outputs
  "estimatr",       # Robust SEs and other estimators
  "DeclareDesign",  # Research design simulation
  "optmatch",       # Optimal matching
  "RItools",        # Randomization inference tools
  "coin",           # Conditional inference
  "boot",           # Bootstrap methods
  "sandwich",       # Robust covariance matrices
  "lmtest"          # Linear model testing
))
```

### Downloading Data

All data files are available at:

**https://github.com/[your-repo]/statistics-textbook/data**

You can:

1. Clone the entire repository
2. Download individual data files
3. Load some directly from R using URLs

### Getting Help

If you encounter problems:

1. **Check the FAQ** (Appendix X)
2. **Search existing issues** on GitHub
3. **Post a new issue** with:
   - What you tried
   - What happened
   - What you expected
   - Your R version and OS
   - A minimal reproducible example

## Accessibility

We strive to make this book accessible to all learners. If you encounter barriers:

- **Vision**: All figures include alt text. Code can be enlarged in RStudio.
- **Color**: Plots avoid relying solely on color to convey information.
- **Mathematics**: Key equations are explained in words, not just symbols.
- **Language**: We avoid jargon when possible and define technical terms.

If you need accommodations not currently provided, please contact us.

## License and Sharing

This book is licensed under **Creative Commons Attribution-NonCommercial-ShareAlike 4.0**.

You can:

- Share it freely
- Adapt it for your own teaching
- Create derivative works

You must:

- Give credit
- Use the same license
- Not use commercially without permission

All code is licensed under MIT License (use freely, even commercially).

## Contributing

Found a typo? Have a better explanation? Created a great extension exercise?

We welcome contributions:

1. Fork the repository
2. Make your changes
3. Submit a pull request

Even small contributions (typo fixes, clarifications) are valuable.

---

Now that you know how to use the book, let's begin exploring statistics!
