---
title: 'Exploration 1: Description in One-Dimension'
author: 'Jake Bowers'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
execute:
  eval: false
  echo: true
format:
  html:
    code-fold: true
    method: mathjax
    url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
  pdf:
    number-sections: true
    colorlinks: true
    cite-method: biblatex
    keep-tex: true
    monofontoptions: "Scale=0.7"
    include-in-header:
      text: |
        \usepackage{bm}
        \usepackage{tikz}
        \usepackage{tikz-cd}
        \usetikzlibrary{arrows,automata,positioning,trees,decorations.markings}
        \usepackage{tikz-qtree}
        \usepackage{amsmath, booktabs, caption, longtable,listings,fancyvrb,fvextra}
        \DeclareOldFontCommand{\bf}{\normalfont\bfseries}{\mathbf}
        \newfontfamily\grouptwofont[Ligatures=TeX]{Fira Sans}
fontsize: 10pt
geometry: margin=1in
graphics: yes
# bibliography: classbib.bib  # Using book-level references.bib instead
biblio-style: "authoryear-comp,natbib"
---

<!-- \input{mytexsymbols} -->

```{r setup, echo=FALSE, results=FALSE, include=FALSE, cache=FALSE}
# Note: qmd_setup.R will be created when populating from course materials
# library(here)
# source(here("qmd_setup.R"))
library(dplyr)
# library(robustbase)
# library(psych)
# library(robustHD)
```


This document is called an "exploration" rather than a "problem set" or
"homework" because the point is to use the discussion, code, results, questions
here to learn something new, practice thinking about statistical concepts,
data, and code and to raise new questions that you'd like to discuss in class.
Some of the code and output might be visible in a pdf file and some might not.
The code has errors that you will need to fix and understand. You should work
with the source document itself as you answer the questions posed by me (the
author) as well as the questions posed by fictional actors who I invent in the
hopes of making these explorations fun. I also expect you to explain the code,
and otherwise explore the data and the questions arising from their discussion.

I expect that:
- you will make an attempt to answer every question posed here
- you will read the pieces that are recommended. You should be able to answer
  questions about what the reading said.
- you will make an attempt to explain to yourself every line of code written in
  this document, 
- you will run all of the code yourself, 
- that you will write some new code of your own as you work on answering the
  questions. (For example, you might look for alternative ways to summarize
  "location" or "spread" other than those used here. You might try different
  arguments to the same functions used here. Or you might find new functions or
  write your own. Or you might try using these functions on different data. You
  might have conversations with an AI to help you understand what is going on
  here and to get suggestions for alternatives with their associated advantages
  and disadvantages.)

Useful reading about description:

 - @wilcox2012introduction, Chap 1--3
 - @kaplan2012ism, Chap 2--3

\bigskip

"Brexit! UKIP! ISIL! Taliban! COVID!" When your old friend calls, she seems to
be yelling. Once she calms down, she explains: "I am in charge of Improving
Civic Society programs for the United Nations, and have been asked to step in
to help out at the UK Office of Social Capital." After you congratulate her on
what appears to be a promotion she continues. "The thing is that over here in
the UK, they are really big on numbers. I asked my staff for a simple report on
the status of civic society in the UK before all of the recent unrest happened
there, say, in 2005, before the London Bombings. They responded with numbers.
When I asked them to explain, I found their desks empty, their chairs knocked
over, their computers smashed, but their coffee cups still warm and untouched."
You ask her about her own safety and she responds. "This is all within
operational parameters. No worries. My problem is that I need to report to the
high command, I mean, my manager, and I don't know what the right answer is.
Now I don't even have numbers. Please help. Can we hop on a Zoom call?"

She does not enable video in the Zoom call. However, she begins sending you
some code. "Here is what I have in terms of output. Can you explain to me what
is going on with each line of code? I mean shouldn't the total number of hours
people spend helping each other be something other than `NA`? How do you figure
out what is going on when you make errors or find something confusing in `R`?"


```{r}
load(url("http://jakebowers.org/Data/ho05.rda"))
table(ho05$postbomb, useNA = "ifany")
wrkdat <- ho05 %>% filter(postbomb == 0)
sum(wrkdat$hlphrs)
```

She continues, "And then I have this from a previous meeting where they talked
about `codebooks` but I don't think these were the ordinary kind of encrypted
communication behind enemy lines it seems like a weird codebook to me."

```
### CODEBOOK
postbomb: 1=interviewed after the bombing, 0=interviewed before the bombing

grphrs: 6.1.1 Which of the following groups, clubs or organisations
  have you been involved with during the last 12 months? That's anything
  you've taken part in, supported, or that you've helped in any way, either
  on your own or with others. Please exclude giving money and anything that
  was a requirement of your job.

  6.1.2 In the last 12 months have you given unpaid help to any groups, clubs or
  organisations in any of the following ways?

  6.1.5 Approximately how many hours have you spent helping this/these group(s),
  club(s) or organisation(s) in the past 4 weeks?

infhrs: In the last 12 months have you done any of the following things,
  unpaid, for someone who was not a relative?

  This is any unpaid help you, as an individual, may have given to other people,
  that is apart from any help given through a group, club or organisation. This
  could be help for a friend, neighbour or someone else but not a relative.

  6.4.4 Now just thinking about the past 4 weeks. Approximately how many hours
  have you spent doing this kind of thing/these kind of things in the past 4
  weeks?

hlphrs: grphrs+infhrs
```

She asks, "What is the best way describe how civic life was going before the
bombings in 2005 (let alone before all of the other disruptions occurred in the
UK)? What is the right answer?"

Later, after you had worked on this a bit she calls back, "Hey. Thanks so much
for helping! I just found this code and thought it might be useful although it
has some errors. What do you think? Can you tell me what this means? Does it
help me get the right answer about how much time people in the UK were devoting
to helping each other and/or supporting groups? Why are there so many ways to
descibe a single variable anyway? What is the point? Which approaches do you
prefer (from those below and others)? Also, are there any plots that would help
me tell the right story about this variable? What might they be? What do they
look like?"

```{r deltaforce1robustdescription, results='hide'}
mean(wrkdat$hlphrs, trim = .1, na.rm = TRUE)
hlp_vec <- unlist(wrkdat %>% dplyr::select(hlphrs) %>% filter(!is.na(hlphrs)))
mean(hlp_vec, trim = .1, na.rm = FALSE)
winsor.mean(hlp_vec)
mean(winsorize(hlp_vec))

one_step_M_est <- function(x) {
  ## Following Section 2.1 in http://www.psychology.mcmaster.ca/bennett/boot09/rt2.pdf
  madn <- mad(x, constant = 1.4826) ## 1/1.4826 = .6745
  M <- (abs(x - median(x)) / madn) > 1.28
  U <- sum(x > M)
  L <- sum(x < M)
  B <- length(x) - U - L
  n <- length(x)
  mux <- (1.28 * madn * (U - L) + B) / (n - L - U)
  return(mux)
}

least_squares_loss <- function(x, the_var) {
  ssr <- sum((the_var - x)^2)
  return(ssr)
}

onestepMest(hlp_vec)
huberM_results <- huberM(hlp_vec)
print(huberM_results$mu)
print(huberM_results$s)

ls_location <- optimize(
  f = least_squares_loss,
  the_var = hlp_vec,
  interval = range(hlp_vec)
)

ls_location

l1_loss <- function(x, the_var) {
  ssr <- sum(abs(the_var - x))
  return(ssr)
}

l1_location <- optimize(
  f = l1_loss,
  the_var = hlp_vec,
  interval = range(hlp_vec)
)

l1_location

fivenum(hlp_vec)
quantile(hlp_vec, seq(0, 1, .1))
summary(hlp_vec)
sd(hlp_vec)
mad(hlp_vec)
```

<!-- see also https://dornsife.usc.edu/labs/rwilcox/software/ and WRS:::mest  https://dornsife.usc.edu/assets/sites/239/docs/WRS2.pdf and the MASS library--->

# References

- Rand R Wilcox (2012). Introduction to robust estimation and hypothesistesting. Academic Press, Chap 1–3

- Daniel Kaplan (2012). Statistical Modeling A Fresh Approach. Second. Macalester College, St. Paul, MN: Daniel Kaplan, Chap 2–3
